---
title: "Dental Arcade: Differential abundance"
author: "Zandra Fagernaes"
output: html_notebook
---

This notebook goes through the differential abundance analysis for the Dental Arcade project.

```{r}
library(janitor)
library(tidyverse)
library(ggpubr)
```

The software Songbird is used for differential abundance studies.Input data is a species level OTU table from a MALT run against the custom RefSeq database. Contaminants (blanks and bones) are cleaned out, and only taxa present in all four individuals are included. Blanks, bones, and occlusal samples are excluded. Another dataset is made where occlusal samples are included, to study the effect of disease on differential abundance.

```{r}
# Import raw data
data <- read.delim("<path_to>/DA_MALT_cRefSeq_species_summarized_20200616.txt")
colnames(data)[1] <- "Species"

# Metadata
meta <- read.delim("<path_to>/da_metadata_new.txt")

# Fix sample names
colnames(data) <- gsub(".SG1.1.2_S0_L001_R1_001.fastq.combined.prefixed.fastq.extractunmapped.bam", 
                           "", colnames(data))
colnames(data) <- gsub(".SG1.1_S0_L001_R1_001.fastq.combined.prefixed.fastq.extractunmapped.bam",
                           "", colnames(data))

# Remove bones, occlusal samples and blanks
data <- data %>%
  select(-c("EXB037.A2101","EXB037.A2201","EXB037.A2301","EXB037.A2401","EXB037.A2501", "EXB037.A3301",
            "LIB030.A0107","LIB030.A0108","LIB030.A0109","LIB030.A0110","LIB030.A0111", "LIB030.A0115",
            "CDM045.Z0101", "CDM047.Z0101", "CDM053.J0101", "CDM054.A0101",
            #"CDM048.E0101", "CDM048.O0101", "CDM049.E0101", "CDM049.I0101"
            ))

# Import contaminants
cont <- read.delim("<path_to>/DA_decontam_cRefSeq_species_20210408.txt")

# Remove contaminants
data_clean <- anti_join(data, cont)

# Remove genera with no counts
data_clean$sum <- rowSums(data_clean[2:88])
data_clean <- data_clean %>%
  filter(!sum==0) %>%
  select(-sum)

# Transpose
data_clean_t <- data_clean %>%
  gather(sample, count, 2:88) %>%
  spread(Species, count)

# Add column with individual
ind <- meta %>%
  select(LibraryID, Individual)
data_clean_t_ind <- left_join(data_clean_t, ind, by=c("sample"="LibraryID"))

# Turn into long format and remove zero counts
data_long <- data_clean_t_ind %>%
  gather(species, count, 2:275) %>%
  filter(count > 0)

# Count how many samples each species is found in per individual
data_prevalence <- data_long %>%
  select(species, Individual, sample) %>%
  group_by(species, Individual) %>%
  summarize(prevalence = n())

# Filter for species present in >3 samples per individual
data_prevalence <- data_prevalence %>%
  filter(!prevalence<3)

# Filter for species present in all four individuals
data_list <- data_prevalence %>%
  select(species, Individual) %>%
  group_by(species) %>%
  summarize(prevalence=n()) %>%
  filter(prevalence==4) %>%
  select(species)

# Create final OTU table
data_final <- left_join(data_list, data_clean, by=c("species"="Species"))
colnames(data_final)[1] <- "# OTU ID"

# Save OTU table
write.table(data_final, "<path_to>/DA_MALT_cRefSeq_species_cleaned_occlusal_3sample_4ind_20210609.txt", sep="\t", row.names = FALSE, quote=FALSE)
```

# Make plot for Songbird results

The data was run through Songbird with a model including all explanatory variables (also individual), as well as a null model. Different models were tried to find ne with good fit.

```{r}
# Import result dataset
songbird <- read_tsv("<path_to>/all_species_cleaned_3sample_4ind_20210409_bestfit.tsv") %>%
  clean_names()

# Filter out only total weight (scaled), top and bottom ten
weight <- songbird %>%
  dplyr::select("featureid", "total_weight_scaled") %>%
  arrange(total_weight_scaled) %>%
  filter(row_number() %in% c(1:10, 137:146))

# Filter out only jawbone, top and bottom ten
jawbone <- songbird %>%
  dplyr::select("featureid", "jawbone_t_maxilla") %>%
  arrange(jawbone_t_maxilla) %>%
  filter(row_number() %in% c(1:10, 137:146))

# Filter out only toothside (i), top and bottom ten
toothside_i <- songbird %>%
  dplyr::select("featureid", "tooth_side_t_i") %>%
  arrange(tooth_side_t_i) %>%
  filter(row_number() %in% c(1:10, 137:146))

# Filter out only toothside (l), top and bottom ten
toothside_l <- songbird %>%
  dplyr::select("featureid", "tooth_side_t_l") %>%
  arrange(tooth_side_t_l) %>%
  filter(row_number() %in% c(1:10, 137:146))

# Filter out only tooth position, top and bottom ten
toothposition <- songbird %>%
  dplyr::select("featureid", "tooth_position_t_posterior") %>%
  arrange(tooth_position_t_posterior) %>%
  filter(row_number() %in% c(1:10, 137:146))

# For dataset with occlusal, filter out only toothside (o), top and bottom ten
toothside_o <- songbird %>%
  dplyr::select("featureid", "tooth_side_t_o") %>%
  arrange(tooth_side_t_o) %>%
  filter(row_number() %in% c(1:10, 137:146))
```

And then we will create a divergent bar chart where all OTUs are labelled.

```{r}
# Weight
ws <- ggplot(weight, aes(x = reorder(featureid, total_weight_scaled), y = total_weight_scaled))+
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "") +
  ggtitle("Total weight (scaled)") +
  theme_classic()

# Jawbone
jaw <- ggplot(jawbone, aes(x = reorder(featureid, jawbone_t_maxilla), y = jawbone_t_maxilla))+
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "") +
  ggtitle("Jawbone (maxilla)") +
  theme_classic()

# Toothside i
tsi <- ggplot(toothside_i, aes(x = reorder(featureid, tooth_side_t_i), y = tooth_side_t_i))+
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "") +
  ggtitle("Tooth surface (interproximal)") +
  theme_classic()

# Toothside o
tsl <- ggplot(toothside_o, aes(x = reorder(featureid, tooth_side_t_o), y = tooth_side_t_o))+
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "") +
  ggtitle("Tooth surface (occlusal)") +
  theme_classic()

# Tooth position
tp <- ggplot(toothposition, aes(x = reorder(featureid, tooth_position_t_posterior), y = tooth_position_t_posterior))+
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "", y = "") +
  ggtitle("Tooth position (posterior)") +
  theme_classic()

ggarrange(ws, jaw, tsi, tsl, tp, nrow=3, ncol=2)
```

# Dataset for 3D mapping

For 3D mapping, we want to summarize the 10 most divergent taxa for each variable we present in the paper, and then average these values across individual. 

```{r}
# Normalize counts
otu_norm <- as.data.frame(scale(raw_data[2:84], center=FALSE,
                                 scale=colSums(raw_data[2:84])))
otu_norm$otu <- raw_data$Species

## Below, repeat for each variable

# Select subset of genera
otu_var <- left_join(weight, otu_norm, by=c("featureid"="otu")) %>%
  select(-total_weight_scaled)

# Transpose
otu_var_t <- otu_var %>%
  gather(sample, count, 2:84) %>%
  spread(featureid, count)

# Fix names
otu_var_t <- otu_var_t %>%
  clean_names()
  
# Sum anterior taxa per individual - tooth position
otu_toothposition_sum <- otu_toothposition_t %>% 
    rowwise() %>% 
    mutate(ind_anterior = sum(c(methanobrevibacter_olleyae, actinomyces_radicidentis, actinomyces_urogenitalis, lachnoanaerobaculum_sp_obrc5_5, actinomyces_sp_oral_taxon_170, eubacterium_sulci, streptococcus_sanguinis, eubacterium_brachy, peptostreptococcus_anaerobius, actinomyces_slackii))) %>%
  mutate(ind_posterior = sum(c(olsenella_sp_hmsc062g07, olsenella_uli, methanobrevibacter_smithii, olsenella_umbonata, olsenella_phocaeensis, bacteroides_pyogenes, dialister_invisus, atopobium_sp_oral_taxon_810, olsenella_sp_kh2p3, methanobrevibacter_wolinii))) %>%
  select(sample, ind_anterior, ind_posterior)

# Sum anterior taxa per individual - interproximal
otu_var_sum <- otu_var_t %>% 
    rowwise() %>% 
    mutate(ind_other = sum(c(actinobaculum_sp_oral_taxon_183, fusobacterium_sp_oral_taxon_203, treponema_sp_omz_838, treponema_medium, treponema_lecithinolyticum, johnsonella_ignava, bacteroidetes_oral_taxon_274, centipeda_periodontii, actinomyces_gerencseriae, eubacterium_saphenum))) %>%
  mutate(ind_ip = sum(c(actinomyces_sp_oral_taxon_170, olsenella_umbonata, methanobrevibacter_sp_ye315, bacteroides_pyogenes, methanobrevibacter_millerae, olsenella_sp_kh2p3, methanobrevibacter_oralis, methanobrevibacter_smithii, methanobrevibacter_wolinii, methanobrevibacter_olleyae))) %>%
  select(sample, ind_other, ind_ip)

# Sum anterior taxa per individual - weight
otu_var_sum <- otu_var_t %>% 
    rowwise() %>% 
    mutate(ind_low = sum(c(fusobacterium_sp_oral_taxon_203, actinobaculum_sp_oral_taxon_183, schaalia_meyeri, treponema_lecithinolyticum, actinomyces_gerencseriae, actinomyces_sp_hmsc062g12, eubacterium_saphenum, desulfomicrobium_orale, schaalia_georgiae, filifactor_alocis))) %>%
  mutate(ind_high = sum(c(gordonibacter_pamelaeae, capnocytophaga_ochracea, atopobium_sp_oral_taxon_810, capnocytophaga_sp_oral_taxon_332, capnocytophaga_sp_oral_taxon_863, capnocytophaga_sp_oral_taxon_326, capnocytophaga_sp_ch_dc_os43, lautropia_mirabilis, eikenella_corrodens, eikenella_sp_hmsc061c02))) %>%
  select(sample, ind_low, ind_high)

# Add metadata
otu_var_sum_meta <- left_join(meta, otu_var_sum, by=c("LibraryID"="sample")) %>%
  filter(Type == "Calculus") %>%
  filter(!ToothSide == "O")

# Summarize per location
otu_var_mapping <- otu_var_sum_meta %>%
  select(ToothNr, ToothSide, ToothClass, ind_low, ind_high) %>%
  group_by(ToothNr, ToothSide, ToothClass) %>%
  summarize(low_mass_taxa = mean(ind_low), 
            high_mass_taxa = mean(ind_high))

# Save
write.table(otu_var_mapping ,"<path_to>/mass_species_20210610.txt", row.names = FALSE, quote=FALSE, sep="\t")
```








