---
title: "DA grapevine reads"
author: "Zandra Fagernas"
output: html_notebook
---

During an initial screening of a subset of the dataset, it was noticed that the samples have quite many reads from grapevine (Vitis vinifera). However, the bone samples also have a lot of grapevine reads - could they be environmental and not dietary? I mapped the human-removed reads (i.e. the data was mapped to the human reference genome, and all human reads were) to the Vitis vinifera reference genome using EAGER1, with mapping quality 37, and damage profiling turned on.

```{r message=FALSE, warning=FALSE}
library(readxl)
library(ggpubr)
library(lmerTest)
library(lme4)
library(janitor)
library(tidyverse)
library(ggeffects)
```


```{r}
# Import data (EAGER report)
grape_raw <- read_csv("<path_to>/Report_grapevine_mapping_eager1_20210120.csv") %>%
  clean_names() %>%
  dplyr::select(sample_name, number_reads_after_c_m_prior_mapping, mapped_reads_after_rm_dup, endogenous_dna_qf_percent, cluster_factor, dmg_1st_base_3, dmg_2nd_base_3, median_fragment_length)

# Import metadata
meta <- read.delim("<path_to>/DA_metadata_new.txt")

# Combine with metadata
grape_meta <- left_join(grape_raw, meta, by=c("sample_name"="LibraryID")) %>%
  filter(!Type == "Blank")
```

Let's plot the percentage grapevine reads per sample. 

```{r}
# Set colours for individual
my_colors = c("grey", "#C70E7B", "#A6E000", "#6C6C9D", "#1BB6AF")
names(my_colors) = grape_meta$Individual %>% unique %>% sort

# Change x to the variable you want to investigate
endo_grape <- ggplot(grape_meta, aes(x=Type, y=endogenous_dna_qf_percent)) +
  stat_boxplot(geom='errorbar') +
  geom_boxplot() +
  ylab("Percent grapevine reads") +
  xlab("") +
  scale_color_manual(values=my_colors) +
  geom_point(aes(col=Individual)) +
  theme_minimal()
```

Let's also take a look at the damage plots, from the output from DamageProfiler.

```{bash eval=FALSE}
## Code block to concatenate Damageprofiler freq files while adding a column for sample name as a third column.
## Should be ran from the eager1 output directory where all sample subdirectories are. 

## Create header
echo -e "pos\t3pG>A\tsample" > output_file.txt
## For each freq file, read line-by-line omitting the header line. Infer sample name from file path, and add that name as a third column. Then save all the lines form all the files into "output_file.txt"
for file in */7-DnaDamage/*/3pGtoA_freq.txt ; do
  while read line; do
    sample_name=$(echo ${file} | cut -d '/' -f 1)
    echo -e "${line}\t${sample_name}"
  done < <(tail -n +2 ${file})
done >> output_file.txt
```

```{r}
# Load raw data
grape_damage <- read.delim("<path_to>/grapevine_damage_5pCtoT_20210120.txt")

# Load metadata
meta <- read.delim("<path_to>/DA_metadata_new.txt")

# Add metadata
grape_damage_meta <- left_join(grape_damage, meta, by=c("sample"="LibraryID"))

# Set colours for individual
my_colors = c("grey", "#C70E7B", "#A6E000", "#6C6C9D", "#1BB6AF")
names(my_colors) = grape_damage_meta$Individual %>% unique %>% sort

# Plot
damage_fig <- ggplot(grape_damage_meta, aes(x=pos, y=X5pC.T, group=sample, colour=Individual)) +
  geom_line(size=1) +
  ylab("5' C>T") +
  xlab("Position") +
  facet_wrap(~Type, scales = "free") +
  scale_color_manual(values=my_colors) +
  theme_minimal()
```

This is quite messy, so let's subset to only individuals with >500 reads mapped to grapevine (after quality filtering and duplicate removal), see reasoning in Mann et al. 2020. 

```{r}
# Get list of sample IDs with >500 reads
grape_500 <- grape_meta %>%
  filter(mapped_reads_after_rm_dup > 500) %>%
  dplyr::select(sample_name)

# Subset damage dataset
grape_damage_500 <- left_join(grape_500, grape_damage_meta, by=c("sample_name" = "sample"))

# Plot
damage_fig_500 <- ggplot(grape_damage_500, aes(x=pos, y=X5pC.T, group=sample_name, colour=Individual)) +
  geom_line(size=1) +
  ylab("5' C>T") +
  xlab("Position (bp)") +
  facet_wrap(~Type) +
  scale_color_manual(values=my_colors) +
  theme_minimal()

ggarrange(endo_grape, damage_fig_500, ncol=2, labels = c("A", "B"), common.legend = TRUE, legend = "right", widths = c(2,3))
```

